---
title: 'Reproducible Research: Course Project 1'
author: "A Suarez-Pierre"
date: "June 16, 2016"
output: html_document
---

```{r echo=F}
options(warn=-1) #avoids warnings messages
```

### 1. Load necessary packages and set working directory
```{r message=FALSE}
library(ggplot2)
library(dplyr)
library(lubridate)
library(gridExtra)


setwd("/Users/asuarez/Desktop/ReproducibleResearch")
```


### 2. Read the dataset and process the data
```{r}
df = read.csv("activity.csv", stringsAsFactors=FALSE)
```



### 3. Histogram of the total number of steps taken each day
```{r}
df = group_by(df, date)
statsByDay = summarise(df, steps=sum(steps, na.rm=TRUE))

ggplot(data=statsByDay, aes(x=date, y=steps)) + geom_bar(stat="identity", fill="blue") + 
      theme_bw() + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
      labs(title="Total Steps Taken Each Day", x="Day", y="Total Count")
```


### 4. Mean and median number of steps taken each day
*Pardon the long list*
```{r}
statsByDay = summarise(df, 
                       mean=round(mean(steps, na.rm=TRUE),2), 
                       median=median(steps, na.rm=TRUE))

print(statsByDay, n=62)
```


### 5. Times series plot of the average number of steps taken 
```{r}
ggplot(data=statsByDay, aes(x=date, y=mean, group=1)) + geom_point(col="blue") + 
      geom_line(col="blue") + theme_bw() + 
      theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
      labs(title="Mean Steps Taken Each Day", x="Day", y="Average")
```


### 6. The 5-min interval that, on average, contains the maximum number of steps
```{r}
df = group_by(df, interval)

statsByInt = summarise(df, mean=round(mean(steps, na.rm=TRUE), 2))
statsByInt = arrange(statsByInt, desc(mean))

max = (statsByInt$interval[1])
print(max)
```

The interval with the max number of steps on average is `r max`.


### 7. Histogram of the number of steps taken each day after missing values are imputed
```{r}
df = group_by(df, date)
statsByDay = summarise(df, steps=sum(steps, na.rm=TRUE))

missing = statsByDay$steps==0
missing = c(FALSE, missing) # array containing days after missing data

dayAfter = statsByDay[missing,]
dayAfter = dayAfter[1:7,] #removes empty row at the bottom
```

Now we print the histogram
```{r}
ggplot(data=dayAfter, aes(x=date, y=steps)) + geom_bar(stat="identity", fill="blue") + 
      theme_bw() + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
      labs(title="Total Steps Taken Each Day", x="Day", y="Total Count")
```


### 8. Panel plot comparing the average number of steps taken per 5-min interval across weekdays and weekends
```{r}
df$date = ymd(df$date)

# Id which are weekdays 
df = mutate(df, weekday=ifelse(wday(date)==1 | wday(date)==7, 0, 1)) 


# Averages by interval 
df = group_by(df, weekday, interval)
statsByInt = summarise(df, mean=round(mean(steps, na.rm=TRUE), 2))


# Separating into 2 dataframes depending on weekday Vs weekend
wkdays = filter(statsByInt, weekday==1)[,2:3]
wkends = filter(statsByInt, weekday==0)[,2:3]


# Generating plots
plot1 = ggplot(data=wkdays, aes(x=interval, y=mean)) + geom_line(col="blue") + theme_bw() + 
        labs(title="Weekdays: Avg Steps By Time of Day", x="Time", y="Average")
plot2 = ggplot(data=wkends, aes(x=interval, y=mean)) + geom_line(col="blue") + theme_bw() + 
        labs(title="Weekends: Avg Steps By Time of Day", x="Time", y="Average")

grid.arrange(plot1, plot2, nrow=2)
```

